---
title: "Wright-Fisher simulations in R"
author: "Emil Malta"
date: "2025-01-03"
toc: true
description: "This post covers a few ways to implement genetic drift simulations in R."
bluesky-comments:
  mute-patterns:
    - "📌"
  mute-users:
    - "did:plc:1234abcd"
  filter-empty-replies: true
  n-show-init: 3
  n-show-more: 2
draft: true
---

## Genetic drift

## tl;dr

## Naive approach

```{r}
wf_sim <- function(p = .5, n = 50, t = 50) {
  freqs <- c(p)
  for(i in 2:t) {
    freqs <- c(freqs, rbinom(1, size = 2*n, prob = freqs[i - 1]) / (2*n))
  }
  freqs
}

plot(
  wf_sim(), type = "l", ylim = c(0,1), 
  main = "Wright-Fisher simulation of genetic drift"
)
```

```{r}
plot(
  wf_sim(), type = "l", ylim = c(0,1), 
  main = "Wright-Fisher simulation of genetic drift"
)

for(i in 1:100) points(wf_sim(), type = "l")

```

## Preallocate your vectors


```{r}
wf_sim_vectorized <- function(p = .5, n = 50, t = 50) {
  freqs <- vector(mode = "numeric", length = t)
  freqs[1] <- p
  for(i in 2:t) {
    freqs[i] <- rbinom(1, size = 2*n, prob = freqs[i - 1]) / (2*n)
  }
  freqs
}
```
{{< lipsum 1 >}}

```{r}
bench::mark(
  wf_sim(t = 5000), 
  wf_sim_vectorized(t = 5000), 
  check = FALSE
)
```

## For loops are fine

```{r}
wf_sim_tidy <- function(p = .5, n = 50, t = 50) {
  purrr::accumulate(
    .x = vector(mode = "numeric", length = t),
    .f = ~ rbinom(1, 2 * n, .x)/(2 * n), 
    .init = p
  )
}
```


```{r}
bench::mark(
  wf_sim(t = 5000),
  wf_sim_vectorized(t = 5000), 
  wf_sim_tidy(t = 5000),
  check = FALSE
)
```


```{r}
library(tidyverse)
crossing(sim = 1:100, p_init = c(.1, .5, .9), n = c(5, 50, 500)) %>% 
  group_by(sim, p_init, n) %>% 
  reframe(
    p = wf_sim_vectorized(p = p_init, n = n),
    t = seq_along(p)
  ) -> df

df %>% 
  ggplot(aes(x = t, y = p, group = sim)) +
  geom_line(alpha = .2) +
  facet_grid(paste("p_init: ", p_init) ~ paste("n: ", n)) +
  scale_y_continuous(labels = scales::percent) 
```

```{r}
wf_sim_mutation <- function(p = .5, n = 50, t = 50, mut_from = 0, mut_to = 0){
  freqs <- vector(mode = "numeric", length = t)
  freqs[1] <- p
  
  for(i in 2:t) {
    p_mut <- (1 - mut_from) * freqs[i - 1] + mut_to * (1 - freqs[i - 1])
    freqs[i] <- rbinom(1, 2 * n, p_mut) / (2 * n)
  }
  freqs
}
```

```{r}
crossing(sim = 1:100, p_init = c(.01, .5, .99), n = c(50, 500, 5000)) %>% 
  group_by(sim, p_init, n) %>% 
  reframe(
    p = wf_sim_mutation(p = p_init, n = n, mut_to = .06, mut_from = .06),
    t = seq_along(p)
  ) -> df

df %>% 
  ggplot(aes(x = t, y = p, group = sim)) +
  geom_line(alpha = .2) +
  facet_grid(paste("p_init: ", p_init) ~ paste("n: ", n)) +
  scale_y_continuous(labels = scales::percent)
```

```{r}
df %>% 
  filter(t == max(t)) %>% 
  ggplot(aes(y = p)) +
  geom_histogram(color = "white") +
  facet_grid(paste("p_init: ", p_init) ~ paste("n: ", n)) +
  scale_y_continuous(labels = scales::percent) 
```


---
title: "Wright-Fisher simulations in R"
author: "Emil Malta"
date: "2025-01-03"
toc: true
description: "This post covers a few ways to implement genetic drift simulations in R."
bluesky-comments:
  mute-patterns:
    - "📌"
  mute-users:
    - "did:plc:1234abcd"
  filter-empty-replies: true
  n-show-init: 3
  n-show-more: 2
draft: true
---

## Genetic drift

## tl;dr

## Naive approach

```{r}
wf_sim <- function(p = .5, n = 50, t = 50) {
  freqs <- c(p)
  for(i in 2:t) {
    freqs <- c(freqs, rbinom(1, size = 2*n, prob = freqs[i - 1]) / (2*n))
  }
  freqs
}

plot(
  wf_sim(), type = "l", ylim = c(0,1), 
  main = "Wright-Fisher simulation of genetic drift"
)
```

```{r}
plot(
  wf_sim(), type = "l", ylim = c(0,1), 
  main = "Wright-Fisher simulation of genetic drift"
)

for(i in 1:100) points(wf_sim(), type = "l")

```

## Preallocate your vectors


```{r}
wf_sim_vectorized <- function(p = .5, n = 50, t = 50) {
  freqs <- vector(mode = "numeric", length = t)
  freqs[1] <- p
  for(i in 2:t) {
    freqs[i] <- rbinom(1, size = 2*n, prob = freqs[i - 1]) / (2*n)
  }
  freqs
}
```
{{< lipsum 1 >}}

```{r}
bench::mark(
  wf_sim(t = 5000), 
  wf_sim_vectorized(t = 5000), 
  check = FALSE
)
```

## For loops are fine

```{r}
wf_sim_tidy <- function(p = .5, n = 50, t = 50) {
  purrr::accumulate(
    .x = vector(mode = "numeric", length = t),
    .f = ~ rbinom(1, 2 * n, .x)/(2 * n), 
    .init = p
  )
}
```


```{r}
bench::mark(
  wf_sim(t = 5000),
  wf_sim_vectorized(t = 5000), 
  wf_sim_tidy(t = 5000),
  check = FALSE
)
```


```{r}
library(tidyverse)
crossing(sim = 1:100, p_init = c(.1, .5, .9), n = c(5, 50, 500)) %>% 
  group_by(sim, p_init, n) %>% 
  reframe(
    p = wf_sim_vectorized(p = p_init, n = n),
    t = seq_along(p)
  ) -> df

df %>% 
  ggplot(aes(x = t, y = p, group = sim)) +
  geom_line(alpha = .2) +
  facet_grid(paste("p_init: ", p_init) ~ paste("n: ", n)) +
  scale_y_continuous(labels = scales::percent) 
```

```{r}
wf_sim_selmut <- function(p = .5, n = 50, t = 50, mut_from = 0, mut_to = 0, 
                            fit_AA = 1, fit_Aa = 1, fit_aa = 1) {

  
  freqs <- vector(mode = "numeric", length = t)
  freqs[1] <- p
  
  
  for(i in 2:t) {
    fit_avg = freqs[i - 1] ^ 2 * fit_AA + 
      2 * freqs[i - 1] * (1 - freqs[i - 1]) * fit_Aa + 
      (1 - freqs[i - 1]) ^ 2 * fit_aa
    
    p_sel = (freqs[i - 1] * (freqs[i - 1] * fit_AA + (1 - freqs[i - 1]) * fit_Aa)) / fit_avg
    p_mut <- (1 - mut_from) * p_sel + (1 - p_sel) * mut_to
    
    freqs[i] <- rbinom(1, 2 * n, p_mut) / (2 * n)
  }
  freqs
}

crossing(sim = 1:100, p_init = c(.1/250, .5, .9), n = c(5, 50, 500)) %>% 
  group_by(sim, p_init, n) %>% 
  reframe(
    p = wf_sim_selmut(p = p_init, n = n, fit_Aa = .5, fit_aa = .25),
    t = seq_along(p)
  ) -> df

df %>% 
  ggplot(aes(x = t, y = p, group = sim)) +
  geom_line(alpha = .2) +
  facet_grid(paste("p_init: ", p_init) ~ paste("n: ", n)) +
  scale_y_continuous(labels = scales::percent) 
```